% ****** Start of file aipsamp.tex ******
%
%   This file is part of the AIP files in the AIP distribution for REVTeX 4.
%   Version 4.1 of REVTeX, October 2009
%
%   Copyright (c) 2009 American Institute of Physics.
%
%   See the AIP README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.1
%
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex  aipsamp
%  2)  bibtex aipsamp
%  3)  latex  aipsamp
%  4)  latex  aipsamp
%
% Use this file as a source of example code for your aip document.
% Use the file aiptemplate.tex as a template for your document.
\documentclass[%
 aip,
 jmp,%
 amsmath,amssymb,
% preprint,%
 reprint,%
%author-year,%
%author-numerical,%
]{revtex4-1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% necessary to get the Polish hook to work, according to
% https://tex.stackexchange.com/questions/75396/how-to-use-polishhook-symbol
% I really don't know what fontenc does, but with T2A it screws up the
% fonts, at least when compiling on my system.
% \usepackage[T1,T2A]{fontenc}
% \usepackage[T1]{fontenc}
% \usepackage[utf8]{inputenc}
% I was told this was necessary, but it messes with things like month
% names and doesn't seem to do anything absolutely necessary here.
% \usepackage[polish,russian]{babel}
% \usepackage[polish]{babel}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[usenames,dvipsnames]{color}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{color}
\usepackage{amssymb}
\usepackage{hyperref}
% \usepackage{caption}
\usepackage[ruled,linesnumbered]{algorithm2e}
%\setlength{\algomargin}{2em}

%%% Not sure which, but one of these packages below busts the table of contents.

%\usepackage{algorithm}
%\usepackage{algorithmic}
%\usepackage{listings}
%\usepackage{multirow}
%\usepackage{parskip}
%\usepackage{tabularx}
%\usepackage{psfrag,fancyhdr,layout,appendix}
%\usepackage{natbib}
%\usepackage{chapterbib}
% \usepackage{verbatim}
% \usepackage{framed}
% \usepackage{amsthm}
% \usepackage{booktabs}
% \usepackage{placeins}
% \usepackage{rotating}
% \usepackage{changebar}
% \usepackage{dcolumn}% Align table columns on decimal point
% \usepackage{multirow}
% \usepackage{longtable}
\usepackage{times}
\usepackage{listings}

\def\outlinecomment#1{\hrule{\color{Blue}$\Rightarrow${\small\em #1}}\hrule}
%\def\comment#1{}
\def\comment#1{{\color{ForestGreen}$\Rightarrow${\small\em #1}}}
%\def\comment#1{}

\def\aonolook#1{{\color{Red}$\Rightarrow$Aono-san, \ifx&#1& %
    can you take a look at this?
    \else
    #1
    \fi}}

\def\ddp#1{{\color{Red}$\Rightarrow$ddp, \ifx&#1& %
    can you take a look at this?
    \else
    #1
    \fi}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% rwd-quantum : helpful macros for writing about quantum theory
%
%
% Inner products bras kets etc 
%
\newcommand{\bra}[1]{%  dirac 'bra'
    \ensuremath{\left\langle #1 \right|}\xspace}

\newcommand{\ket}[1]{%  dirac 'ket'
    \ensuremath{\left|  #1 \right\rangle}\xspace}

\newcommand{\innp}[2]{%  dirac style inner product
    \ensuremath{\langle #1 \mid #2 \rangle}}

\newcommand{\outp}[2]{%  dirac style outer product
    \ensuremath{\left|\left. #1 \rangle\right.\!\langle\left. #2 \right|\right. }}

\newcommand{\norm}[1]{% define command to get spacing right
        \ensuremath{\left| #1 \right| }}

\newcommand{\braket}[2]{\ensuremath{\langle #1|#2 \rangle}}
\newcommand{\ketbra}[2]{\ensuremath{|#1 \rangle \langle #2|}}
\newcommand{\ro}[1]{\ensuremath{|#1 \rangle \langle #1|}}
\newcommand{\av}[1]{\ensuremath{\langle #1 \rangle}}
\newcommand{\real}{\ensuremath{\mathrm{Re}}}
\newcommand{\trace}{\ensuremath{\textsf{Tr}}}

%% use the proper notation for CZ etc
\newcommand{\CZ}{\ensuremath{\wedge Z}\xspace}
\newcommand{\CX}{\ensuremath{\wedge X}\xspace}
\newcommand{\CNot}{\CX}
\newcommand{\cnot}{\textsc{cnot}\xspace}
\newcommand{\ccnot}{\textsc{ccnot}\xspace}

\newcommand{\PsiPlus}{\ensuremath{\ket{\Psi^+}}\xspace}
\newcommand{\PsiMinus}{\ensuremath{\ket{\Psi^-}}\xspace}
\newcommand{\PhiPlus}{\ensuremath{\ket{\Phi^+}}\xspace}
\newcommand{\PhiMinus}{\ensuremath{\ket{\Phi^-}}\xspace}

\newcommand{\sulPisP}{\ensuremath{\bra{\Psi^+}}\xspace}
\newcommand{\suniMisP}{\ensuremath{\bra{\Psi^-}}\xspace}
\newcommand{\sulPihP}{\ensuremath{\bra{\Phi^+}}\xspace}
\newcommand{\suniMihP}{\ensuremath{\bra{\Phi^-}}\xspace}

% useful operator names
% trace
\newcommand{\Tr}{\operatorname{Tr}}
% probability
\newcommand{\Prob}{\operatorname{Pr}}
% projection operator 
% \newcommand{\Proj}{\operatorname{P}}



\newcommand{\myindent}{\hspace*{6pt}}
\newcommand{\myindentc}{\hspace*{15pt}}
\newcommand{\R}{\ensuremath{{\sf R\hspace*{-0.9ex}\rule{0.15ex}
{1.5ex}\hspace*{0.9ex}}}}
% \newcommand{\cnot}{\leavevmode\hbox{\footnotesize{CNOT }}}
% \newcommand{\trace}{\ensuremath{\textsf{Tr}}}
\newcommand{\oth}[3]{\ensuremath{\left( \begin{array}{c} #1 \\ #2 \\ #3
\end{array} \right)}}
\newcommand{\me}{\ensuremath{\mathrm{e}}}
\newcommand{\mi}{\ensuremath{\mathrm{i}}}
\newcommand{\mmlink}{\ensuremath{M\rightarrow M}\xspace}
\newcommand{\mmlinkleft}{\ensuremath{M\leftarrow M}\xspace}
\newcommand{\mimlink}{\ensuremath{M\rightarrow I \leftarrow M}\xspace}
\newcommand{\msmlink}{\ensuremath{M\leftarrow S \rightarrow M}\xspace}

\RequirePackage{ifthen}

\newboolean{ShowComments}
\setboolean{ShowComments}{true}  % change to true to show remaining colorful author comments. 
                                 % change to false to hide all author comments.
\ifthenelse{\boolean{ShowComments}}%
	{
		\newcommand{\ColorComment}[3]{%
				{\colorbox{#1}{\color{White}   \textsf{\textbf{#2}}} \textcolor{#1}{#3}}}%  Colorful box, initials, phrase 
        \newcommand{\TR}[1]{\textcolor{Red}{#1}}
        \newcommand{\TB}[1]{\textcolor{Red}{#1}}
        \newcommand{\TO}[1]{\textcolor{Red}{#1}}
	}%
	{
		\newcommand{\ColorComment}[3]{}%  Do nothing at all
        \newcommand{\TR}[1]{#1}
        \newcommand{\TB}[1]{#1}
        \newcommand{\TO}[1]{#1}
	}%

% new commandshttps://www.overleaf.com/project/60b8db48bd3baa062ef3a510
%\newcommand{\TR}{\textcolor{red}}
%\newcommand{\TB}{\textcolor{blue}}
%\renewcommand\TR[1]{}
%\renewcommand\TB[1]{}
\definecolor{rdvcolor}{rgb}{0,0.5,0}\newcommand{\rdv}[1]{\ColorComment{rdvcolor}{rdv}{#1}}

\begin{document}

%\preprint{AIP/123-QED}

\title[What Crypto]{What Every Quantum Researcher and Engineer Should Know about Classical Cryptography}% Force line breaks with \\
% \thanks{Supported by JSPS and AOARD.}
% \thanks{Supported by AOARD.}

\author{Rodney Van Meter}
%  \altaffiliation[Also at ]{}%Lines break automatically or can be forced with \\
 \email{rdv@sfc.wide.ad.jp.}
\affiliation{
Faculty of Environment and Information Studies, Keio University.%\\This line break forced with \textbackslash\textbackslash
}%

\author{Yoshinori Aono}
%  \altaffiliation[Also at ]{}%Lines break automatically or can be forced with \\
 \email{aono@nict.go.jp.}
\affiliation{
National Institute of Information and Communications Technology.%\\This line break forced with \textbackslash\textbackslash
}%

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified

\begin{abstract}
  The potential impact of Shor's factoring algorithm on public key
  cryptography and the potential of quantum key distribution to
  replace key agreement mechanisms such as Diffie-Hellman are two of
  the first facts that quantum researchers learn, but many researchers
  remain vague on how those mechanisms fit into the overall goal of
  achieving confidentiality, integrity and availability in classical
  communications.  We describe the relationship between cryptography
  and communication protocols, with emphasis on key lengths, rekeying,
  block sizes and known mathematical attacks on various ciphers.  We
  primarily focus on Internet encryption, but also introduce WiFi and
  cellular telephone networks.  This information will help quantum
  researchers understand when quantum computers will affect classical
  cryptographic practice and discover new avenues for applying quantum
  technology to the overall problem of secure communication.
%
%\comment{Valid PACS numbers may be entered using the \verb+\pacs{#1}+ command.}
\end{abstract}

\pacs{Valid PACS appear here}% PACS, the Physics and Astronomy
                             % Classification Scheme.
\keywords{Suggested keywords}%Use showkeys class option if keyword
                              %display desired
\maketitle

\tableofcontents{}
%%% I can't get the ToC to work properly on my installation, complains
%%% about ! Missing number, treated as zero.
%%% and ! Illegal unit of measure (pt inserted).
%%% I think it's coming from \tocleft being undefined,
%%% though it also seems to be having trouble finding
%%% the section numbers. --rdv, 18/1/8
% \tableofcontents

\input{intro}
\input{encryption}
\input{cryptanalysis}
\input{comm-standards}
\input{q-attacks}
\input{post-q}
\input{policy}

\section*{Acknowledgments}

This document began as a very long email to a group researchers,
following up on a question from Renato Renner about cryptographic key
lifetimes at the Network Games, Tropical Topology and Quantum
Communication Workshop in Berlin in 2019.  Thanks to Jens Eisert for
inviting me.

Thanks to Darrell Piper for insight into the early discussions of IKE
and IPsec.

I used to work on a box known as an IPsec gateway, but I was the OS
guy, not the cryptographer; Darrell Piper, Dan Harkins and Dave
Kashtan did most of the crypto work.  I am also a board member of the
WIDE Project, which through its KAME Project created the Racoon
software that was an important early implementation of the key
exchange protocol for IPsec, with much of that work done by Shoichi
Sakane.  Much of what I know about IPsec and other security was just
incidental radiation from them, or gained from the security-oriented
papers we had to read when I did my master's at USC back before the
Englightenment.

\section{References}

\bibliography{rdv.bib,refsaono.bib}

\input{temporary}

\appendix
\onecolumngrid

\input{blockchain}
\input{unsorted-notes}
\input{code-examples}
\input{performance}

\end{document}

Diffie \& Hellman analyzed direct, brute force attacks on DES, and
found them to be feasible, all the way back in 1977.
\url{https://ieeexplore.ieee.org/abstract/document/1646525}


The above I knew off the top of my head, but I'm not a cryptographer
and had more or less taken the need for changing the keys (known as
rekeying) on faith.  So, let's dig a little deeper...there won't be
any \emph{new} knowledge in this, but I hope it helps bring together some
information and organizes it to both save you the googling and
calculating I've done and to provide a clear picture.

Shelby Kimmel on no-go bounds for quantum seals:
\url{https://journals.aps.org/pra/accepted/45074N5dCa413a18c5645123e9b2b8363d15ea1fc}

The Best Thing (TM)

Some of the references I used for this section:

The best source on the current state of the birthday paradox is Bhargavan and Leurent~\cite{bhargavan2016practical}.  Other references include Miessler~\footnote{\url{https://danielmiessler.com/study/birthday_attack/}} and Wolfram Mathworld~\footnote{\url{http://mathworld.wolfram.com/BirthdayAttack.html}}.

1. Howard M. Heys, "A tutorial on linear and differential
cryptanalysis", undated but probably 1999ish?
\url{http://www.engr.mun.ca/~howard/Research/Papers/ldc_tutorial.html}

2. Abdalla and Bellare,
\url{https://link.springer.com/chapter/10.1007/3-540-44448-3_42}
\url{http://cseweb.ucsd.edu/~mihir/papers/rekey.html}
found via
\url{http://cseweb.ucsd.edu/~mihir/}

That paper talks about differential/linear cryptanalysis and about the
birthday paradox, saying block size $k$ needs to be rekeyed every
$2^{k/2}$ blocks.

Bellare et al, A concrete security treatment of symmetric encryption:
analysis of the DES modes of operation
abstract from STOC 1997
https://ieeexplore.ieee.org/abstract/document/646128
full paper at
\url{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.117.4734&rep=rep1&type=pdf}
Focuses on DES, which was in the process of being superseded by AES
even in 1997, but the content of the paper is valuable with respect to
CBC.  I found the paper a tough read when trying to figure out how to
apply the equations.

Most important reference:

\url{https://sweet32.info/}

Following their seminal paper on the technique, Biham \& Shamir
published an entire book on the topic:
Biham \& Shamir, Differential Cryptanalysis of the Data Encryption
Standard, Springer 1993.
\url{http://www.cs.technion.ac.il/~biham/Reports/differential-cryptanalysis-of-the-data-encryption-standard-biham-shamir-authors-latex-version.pdf}

:
\url{https://arxiv.org/abs/quant-ph/9601021}
\url{https://journals.aps.org/pra/abstract/10.1103/PhysRevA.54.2605}
